Estimating duration using survival analysis

### 2023-08-10 cis-perfect-testing.tex
Updates from Overleaf

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or prior specification (see \cref{perf-test:sec:simulation-study}).
However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.

@@

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can be undetected and the infection episode endpoints (start and end times) are known imprecisely.
This chapter develops a method to account for the infrequent testing.
This section describes my approach and its background within the relevant statistical literature.

I analyse the data using survival analysis.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts, \ie becoming detectable; and the terminating event is when the duration ends, \ie becoming no longer detectable.

\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS 
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
The data limitation is due to the infrequent sampling in the CIS study design, meaning that most episodes have only a single positive test.
With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.

The first complication in the analysis is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
Undetected episodes are important to account for because they are systematically shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because individuals are tested rarely compared to the duration of an infection episode, particularly when tested only every 28 days.
However, we do still know the testing schedule of the individual as they are in the CIS cohort.
Undetected events where we have some data on the events in which these events occur is an unusual situtation.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the endpoints of detected infection episodes are known imprecisely.
An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
Censored data is present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiaing and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they were normally applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
Therefore, conditioning on this period overstates the probability of the infection being detected.

Double interval censored and truncated data frequently occur when estimating the survival time of bird nests~\autocite{heiseyABCs}.
In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
In bird nesting studies, the visit times are normally global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.

I perform inference within the Bayesian paradigm for two reasons.
First, the standard methods for estimating the parameter's variance and covariance are poor.
Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
These assumptions may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).

Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling}.
These methods augment the data with the unobserved times of infection, creating a parameter per detected infection.
This would be many parameters for CIS (which has over 4000 in the time period considered in the main analysis of \cref{E-imperf-test}) and therefore computationally expensive.
I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.

@@

In this section, I develop a model accounting for double interval censoring and truncation.
I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
These individuals may have truncated episodes or may not have been infected at all.
However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).

Ghost infections are a mathematical convenience to correct for truncated episodes.
Consider that we draw many episodes independently, from a distribution of infection times and durations with all other characteristics identical (most notably the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs).
This probability space is equivalent to any two random variables which can be over any two of infection time, recovery time, and duration.
Conceptually, we will condition on the total number of episodes drawn (which is unknown) corresponding to the detected episode $i$, $n_i$, then consider the probability of observing exactly one episode and this episode occurring in the interval censored ranges observed (being \emph{admissible}), and then finally marginalise over $n_i$ to form the likelihood contribution for this episode.

Unfortunately, there is no clean interpretation of the number of ghost episodes in terms of real-world quantities; in this sense, it is a pure model parameter.
In reality, there is only one individual who is identical to the individual in which the episode occurs in all relevant ways.
Even within the simulation study (see \cref{perf-test:sec:simulation-study}) these are not infections that are simulated, and understanding the ground truth number of ghost infections is not straightforward.

To formalise this, within the framework of \textcite{heiseyModelling}, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.

To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
Each episode, whether observed or not, must fall into one of these three classes with a probability that is a function of $\theta$.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and
  recovery time within their respective intervals observed.
  $n_{ia} =1$ of these occurred, each with probability
  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
\item
  Truncated episodes, $\Omega_i^C$, which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, $n_{it}$ of these observed, each with probability
  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which do not have an infection and
  recovery time within their respective intervals but would have been
  observed (not truncated). This is all remaining episodes not in the
  previous sets. $n_{iu} = 0$ of these occurred, where they would have
  occurred with probability
  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

@@

Episode $i$ and its ghosts independently occur belong to one of these
classes. Therefore, conditional on $n_i$, the number of episodes in
each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
distributed multinomially. Hence:
\begin{align}
&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
&= n_i p_{ia} p_{it}^{n_i- 1}
\end{align}

The posterior we are interested in is
$p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
where $\theta$ are the parameters governing the survival distribution
and hence allow derivation of the $p$s.
\begin{align}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
\end{align}
where the last line follows by assuming prior independence.

In the remainder of the section, we derive the expressions for $p_{ia}$ and
$p_{it}$ in terms of the data, and analytical solutions for the sum
under different priors.

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
Furthermore, this is the reference prior for this quantity\autocite{heBayesiana}.
Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The above allows us to sample from the posterior without needing to
sample the $n_i$s. This is convenient because there are as many
$n_i$s as episodes observed and hence is computationally costly to
sample. Furthermore, each $n_i$ is a discrete parameter, and hence
cannot be sampled within the current Stan implementation.

@@

Due to the conditional independence of the $n_i$s, we find that:
\begin{align}
&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
&\propto n_i p_{it}^{n_i- 1} p(n_i)
\end{align}
With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
distribution with parameter $1 - p_{it}$. With a negative binomial
prior, we have:
\begin{align}
n_i p_{it}^{n_i- 1} p(n_i)
&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
\end{align}
which is a negative binomial pmf with size parameter $r+1$ and
probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
of this is
\begin{align}
\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
\end{align}

>>>>

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset, however, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence (weakly) informative priors should be used.
Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is weakly or strongly informative (see \cref{perf-test:sec:simulation-study}).
However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.

@@

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
This chapter develops a method to account for the infrequent testing pattern.

I approach the problem through survival analysis, the area of statistics concerned with estimating the distribution of the times between two events.
The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.

\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
Due to the infrequent sampling in the CIS study design, most detected episodes have only a single positive test, and consequently,
there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
However, we do still know the testing schedule of the individual as they are in the CIS cohort.
Undetected events where we have information on missed events is unusual within the survival analysis literature.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the endpoints of detected infection episodes are known imprecisely.
An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
Censored data are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
However, this is not the case here.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
This phase is not representative of the overall study: infections are more likely to be detected during this phase.
Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.

Censored and truncated data frequently occur when estimating the survival time of bird nests, with the censoring taking a variety of forms~\autocite{heiseyABCs}.
In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
In their application, the visit times are global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.

Inference will be performed within the Bayesian paradigm for two reasons.
First, the standard methods for estimating the parameter's variance and covariance are poor.
Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).

Bayesian methods have previously been applied to nest survival settings~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
Therefore, data augmentation applied to the 4000 detected episodes in the main analysis of \cref{E-imperf-test} greatly increases the dimensionality of the problem, making it computationally infeasible.
I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.

@@

In this section, I develop a model accounting for double interval censoring and truncation, modelling only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.

Ghost infections are a mathematical convenience to correct for the undetected episodes which arise through the following thought experiment.
For each detected episode $i$, let $n_i-1$ be the number of undetected episodes, such that $n_i$ is the total number of episodes.
Further, assume that each of these $n_i$ infections episodes were drawn independently from $i$'s distribution of episode beginning and end times and all other characteristics of these infection episodes are identical; most notably, the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs.
If we had detected all $n_i$ episodes, then our sample of detected episodes would be unbiased; however, we did not, and the detected episode is systematically longer than the other $n_i-1$ episodes (as explained in \cref{perf-test:sec:problem}).

In the following method, I consider $n_i$ and the duration of the missing episodes as missing data.
With the full dataset of $n_i$ infection episodes, the problem reduces to double interval censored survival analysis, which has a known likelihood.
Therefore, I will consider the posterior augmented with all the $n_i$s and then integrate over the $n_i$s to form the desired posterior.

\todo[inline]{Maybe move the following paragraph elsewhere}
It is important to note that there is not a correspondence between the number of ghost episodes and the number of infection episodes that were missed within the true cohort.
Within the true cohort, the episodes in which the undetected episodes occurred are different to those in which detected episodes occurred.
In particular, the individuals in which undetected episodes occurred would (on average) have longer spacing between their tests.

The framework of \textcite{heiseyModelling} formalises this thought experiment.
Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.

To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of $\theta$.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and
  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
  $n_{ia} =1$ of these occurred, each with probability
  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
\item
  Truncated episodes, $\Omega_i^C$, which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, $n_{it}$ of these observed, each with probability
  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  observed (not truncated). This is all remaining episodes not in the
  previous sets. $n_{iu} = 0$ of these occurred, where they would have
  occurred with probability
  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

@@

Episode $i$ and its ghosts independently belong to one of these
classes. Therefore, conditional on $n_i$, the number of episodes in
each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
distributed multinomially. Hence:
\begin{align}
&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
&= n_i p_{ia} p_{it}^{n_i- 1}
\end{align}

The posterior we are interested in is
$p(\theta \mid n_{1a} = 1, n_{1u} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
where $\theta$ are the parameters governing the survival distribution
and hence allow derivation of the $p$s.
\begin{align}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i).\label{eqn:sum}
\end{align}

In the remainder of the section, expressions for $p_{ia}$,
$p_{it}$, and analytical solutions for the sum in \cref{eqn:sum} are derived in terms of the observed data, 
under different priors.

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
Furthermore, this is the reference prior for this quantity \autocite{heBayesiana}.
Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The above allows us to sample from the posterior without needing to
sample the $n_i$s. This is convenient because there are as many
$n_i$s as episodes observed and hence is computationally costly to
sample. Furthermore, each $n_i$ is a discrete parameter, and hence
cannot be sampled with widely-used inference algorithms such as Stan's Hamiltonian Monte Carlo (HMC)\todo{ref}.

@@

Due to the conditional independence of the $n_i$s, we find that:
\begin{align}
&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
&\propto n_i p_{it}^{n_i- 1} p(n_i)
\end{align}
With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
distribution with parameter $1 - p_{it}$. With a negative binomial
prior, we have:
\begin{align}
n_i p_{it}^{n_i- 1} p(n_i)
&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
\end{align}
Therefore, by comparing this expression to the pmf of a negative binomial, we find that $n_i - 1$ is distributed negative binomial with size parameter $r+1$ and
probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
of this distribution is
\begin{align}
\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
\end{align}
